---
title: "Fitting linear mixed-effects model using lme4"
output: html_notebook
---

# Aim
To demostrate how to use lme4::lmer() function to fit linear mixed-effects models, and to compare these models. I also wanted to visualize the difference between the different models, and show how to do residual diagnostics automatically and manually. Also, I show a way to create (almost) publication-ready model comparison table.

```{r libraries, include=FALSE}
# lme4 tutorial
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom)
library(modelr)
library(ggfortify)
library(stargazer)
library(gridExtra)
```

We are going to use the built in sleepstudy dataset. It contains reaction times during 10 days of 3-hour-long sleep deprivation. It is a within-subject design, so in theory, it makes sense to use LMM.

```{r}
# Create the data for plotting different random structures
subject_is <- 
    sleepstudy %>% 
    group_by(Subject) %>% 
    do(lm = lm(Reaction ~ Days, data = .) %>% tidy()) %>% 
    unnest() %>% 
    select(Subject, term, estimate) %>% 
    mutate(term = case_when(term == "(Intercept)" ~ "intercept",
                            term == "Days" ~ "slope",
                            TRUE ~ NA_character_)) %>% 
    spread(term, estimate) %>% 
    mutate(fixed_slope = mean(slope),
           fixed_intercept = mean(intercept))
```

# Step 1. Check the data like the observations were independent, i.e. use simple linear regression
```{r, fig.align='center'}
# Create a plot that shows whis logic
ggplot(data = sleepstudy) +
    aes(x = Days, y = Reaction) +
    geom_point() +
    geom_smooth(method = "lm")

# Fit liear regression
sleep.lm <- lm(Reaction ~ Days, data = sleepstudy)
# Check predictor statistics
summary(sleep.lm)
tidy(sleep.lm)
# Check model fit statistics
glance(sleep.lm)
```

Alternatively, we can plot the same regression on each participant, to see how far the fixed intercepts and slopes are from the data points.

```{r, fig.width=9, fig.height=8}
# Plot the fixed intercept and slope model == linear regression
sleepstudy %>% 
    left_join(subject_is) %>% 
    ggplot() +
        aes(x = Days, y = Reaction) +
        geom_abline(aes(intercept = fixed_intercept, slope = fixed_slope), color = "blue", size = 1.2) +
        geom_point() +
        facet_wrap(~Subject)
```

Let's also make a set of residual diagnostic plots.

```{r, fig.height=8, fig.width=9}
# Residual diagnostic plots
autoplot(sleep.lm, which = 1:6)
```

# Step 2. Check if the mixed-model makes more sense
First try a random intercept model. 

The lme4 syntax is very similar to the lm syntax. You can add random terms with + (slope|intercept). If you set the slope to 1, it means it remains fixed.

For a repeated measures design, the random intercept should be the subject (id)
Let's check if the random intercept (but not slope) model is a better fit than the fixed$

```{r, fig.width=9, fig.height=8}
# I want to show the fixed slope, random intercept model
# To do this, I fit an lm to each subject individually (see beginning of script). To get the fixed intercept, I take the average intercept in all models.

# Plot the random intercept model
sleepstudy %>% 
    left_join(subject_is) %>% 
    ggplot() +
        aes(x = Days, y = Reaction) +
        geom_abline(aes(intercept = intercept, slope = fixed_slope), color = "blue", size = 1.2) +
        geom_point() +
        facet_wrap(~Subject)

# Fit the random intercept model
sleep.i <- lmer(Reaction ~ Days + (1|Subject), data = sleepstudy)

summary(sleep.i)
glance(sleep.i)

```
We can see that the model looks better, but the predicted values are sometimes quite off, e.g. with particpant 308 and 335.

## Let's compare the random intercept model with the linear regression model.
The one with the random intercept has a much better fit, although it is more compex.
lme4 by default uses REML for fitting the model, but when we are comparing models, it automatically refits the models in ML, so they are comparable.
```{r}
# Compare random intercept model to fixed intercept model (linear regression)
anova(sleep.i, sleep.lm)
```

# Let's check if we need random slope too, or random intercept is enough
It turns out that the random intercept and slope model is the better fit, so it should be kept as final.
```{r, fig.width=9, fig.height=8}
# First, visualize how the data would look like with random intercept and slope. With ggplot, it is easy, beacause the geom_smooth(method = "lm") function does this automatically, we don't have to use our data from the earlier chunk.
ggplot(sleepstudy) +
    aes(x = Days, y = Reaction) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    facet_wrap(~Subject)
# Fit random intercept and slope model
sleep.is <- lmer(Reaction ~ Days + (Days|Subject), data = sleepstudy)

summary(sleep.is)
glance(sleep.is)
# Compare model to random intercept model
anova(sleep.i, sleep.is)
```

# Residual diadnostics of the linear mixed-model
Sadly, the autoplot is not capable to make the residual diagnostic plots, but we can do it easily using the results of the augment() function.
Despite the three outlier data points, the residuals look rather normal. 
We should accept this model as the final one.

```{r}
# Get the residuals using the broom::augment() function
diag_data <- augment(sleep.is, sleepstudy) %>% 
    mutate(.stdresid = scale(.resid) %>% as.numeric())

# Plot the data together in one object
plot_1 <- ggplot(diag_data) +
    aes(x = .fitted, y = .resid) +
    geom_point() +
    geom_smooth()

plot_2 <- ggplot(diag_data) +
    aes(x = .fitted, y = sqrt(abs(.stdresid))) +
    geom_point() +
    geom_smooth()

plot_3 <- ggplot(diag_data) +
    aes(sample = .stdresid) +
    geom_qq_line(color = "blue") +
    geom_qq()

plot_4 <- ggplot(diag_data) +
    aes(x = .stdresid) +
    geom_histogram(bins = 30)

gridExtra::grid.arrange(plot_1, plot_3, plot_2, plot_4)

```

# Print model comparison using stargazer
We need to refit the mixed-models using the lme4 version of lmer, as stargazer cannot read the lmerTest::lmer() object.

```{r, results = "asis"}
# Refit LMMs in order to show in stargazer
sleep.i.final <- lme4::lmer(Reaction ~ Days + (1|Subject), data = sleepstudy)
sleep.is.final <- lme4::lmer(Reaction ~ Days + (Days|Subject), data = sleepstudy)

stargazer(sleep.lm, sleep.i.final, sleep.is.final, type = "html")
```




